---
title: "AMU workflow example"
author: "Dr. Naomi Kemunto Peter"
date: "2026-06-11"
output: html_document
---

# My role

I independently developed the analytical workflow, including data cleaning, validation procedures, antimicrobial use quantification, statistical modelling, visualization, and reporting.

# Skills demonstated

Data cleaning and validation
Longitudinal data management
Epidemiological analysis
Antimicrobial use quantification
Mixed-effects regression modelling
Data visualization
Reproducible reporting using R Markdown

# Overview
This R Markdown file provides an example analytical workflow from a completed PhD research project on antimicrobial use in semi-intensive poultry systems in Kenya.

The original data are not publicly available because they are subject to research data restrictions. The workflow is provided to demonstrate coding structure,data cleaning procedures, analytical thinking, visualization, and reproducible reporting practices.


# Project objective
 The objective of this analysis was to quantify antimicrobial use in semi-intensive poultry farms and describe temporal patterns of use across production cycles using mass-based antimicrobial use metrics and evaluate association of mortality with antimicrobial dosing practices.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

###Loading the following packages to support data cleaning and analysis

library(tidyverse)
library(epiDisplay)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(reshape)
library(reshape2)
library(data.table)
library(magrittr)
library(lubridate)
library(httr)
library(RColorBrewer)
library(tools)
library(stats)
library(janitor)

```


# Data import

```{r}

# This workflow is an excerpt from a larger analytical pipeline developed during a completed PhD research project.

# Some intermediate datasets and objects are generated in preceding processing steps that are not included in this example repository.

# The original research data are not publicly available due to data-sharing restrictions. A demonstration dataset is referenced here to illustrate the analytical workflow.


AMU_data<-read.csv("AMU_demonstration_data.csv")


```


# Data cleaning and validation


```{r}

###cleaning and validating data

AMU_data1<-AMU_data%>%
  
  ###correct the wrongly entered number of chicken present
  
  dplyr::mutate(
    number_chicken_present=case_when(
     number_chicken_present==598 & 
        date_on_form %in% c("2026-04-15","2026-04-16","2026-04-17","2026-04-18","2026-04-19","2026-04-20","2026-04-21","2026-04-22","2026-04-23","2026-04-24","2026-04-25") & 
        Farm_ID=="A0345" & 
        No_cycles==1~596,
    
    number_chicken_present %in% c(987) & 
      date_on_form %in% c("2026-04-17") & 
      Farm_ID %in% c("A0354") & 
      No_cycles==1~985,
    
    number_chicken_present %in% c(885) & 
      date_on_form %in% c("2026-04-20") & 
      Farm_ID %in% c("A0355") & 
      No_cycles==1~884,
    
    number_chicken_present %in% c(984) & 
      date_on_form %in% c("2026-04-24") & 
      Farm_ID %in% c("A0304") & 
      No_cycles==1~983,
  TRUE~number_chicken_present))%>%
  
  ###correct the wrongly entered number of chicken given medicine
  
   dplyr::mutate(number_chicken_medicine=case_when(
     number_chicken_medicine==771 & 
       Farm_ID=="A0357_2" & 
       date_on_form=="2026-05-23" & 
       No_cycles==1~774,
     
     number_chicken_medicine==218 & 
       Farm_ID=="A0107" & 
       date_on_form %in% c("2026-04-22","2026-04-23","2026-04-24","2026-04-25","2026-04-26") &
       No_cycles==1~699,
     
     number_chicken_medicine==913 & 
       Farm_ID=="A0108_2" & 
       date_on_form %in% c("2026-05-14","2026-05-15","2026-05-16","2026-05-17","2026-05-18") &
       No_cycles==1~1012,
     
     number_chicken_medicine==900 & 
       Farm_ID=="A0101" & 
       date_on_form %in% c("2026-05-24") & 
       No_cycles==1~595,
   TRUE~number_chicken_medicine))%>%
  
   ###correct the wrongly entered number of chicken dead
  
  dplyr::mutate(number_of_dead_chicken=case_when(
    number_of_dead_chicken==80 & 
      Farm_ID=="A01080" & 
      date_on_form %in% c("2026-07-03","2026-07-04","2026-07-05") & 
      No_cycles==1~0, 
    
    number_of_dead_chicken==0 & 
      Farm_ID=="A010_2" & 
      date_on_form %in% c("2026-07-17") & 
      No_cycles==3~5,
    
     number_of_dead_chicken==0 & 
      Farm_ID=="A0108_3" & 
      date_on_form %in% c("2026-07-28") & 
      No_cycles==3~2,
  TRUE~number_of_dead_chicken))%>%
  
   ###correct the wrongly entered number of Day-old-chicken received
  
  dplyr::mutate(number_doc=case_when(
    number_doc==950 & 
      Farm_ID=="A0383" & 
      date_recieved %in% c("2026-08-27") & 
      No_cycles==4~954,
  TRUE~number_doc))%>%
      
  ####remove the duplicated entries
  
dplyr::filter(!(Farm_ID=="A0359" & No_cycles==1 & date_on_form =="2026-05-03" & type_medicine=="antibiotics"))%>%
  
  dplyr::filter(!(Farm_ID=="A0351" & No_cycles==1 & date_on_form =="2026-05-05" & number_chicken_medicine==300))%>%

dplyr::filter(!(Farm_ID=="A0353" & No_cycles==1 & date_on_form =="2026-05-13" & is.na(number_chicken_medicine)))%>%

  distinct()%>%
  
   ###correct the wrongly entered amount of water for medicine dilution
  
  dplyr::mutate(amt_of_water=case_when(
    coversheet.farm_barcode=="A0341" & 
      date_on_form %in%c("2026-07-21","2026-07-23") & 
      name_of_antibiotics=="Colistin" & 
      amt_of_water==70~80,
                                                 
    coversheet.farm_barcode=="A0171" & 
      production_cycles==151 & 
      name_of_antibiotics=="Super skajcycline" & 
      amt_of_water==20~5,
    TRUE~amt_of_water))

```


# Antimicrobial use quantfifaction

```{r}

###quantification of antimicrobial usage using mass-based metrics

####Mass based for all farms per farm per cycle

####data for all farms included in the study
farmCycles7<-farmCycles1


####combine farms data with antimicrobial use data

allfarmCycles<-merge(farmCycles7,daily_AMU,by.x ="coversheet.farm_barcode",by.y = "coversheet.farm_barcode",all.x = TRUE )


allfarmCycles1<-allfarmCycles%>%
  dplyr::select(coversheet.farm_barcode,Farm_label,max_cycles,AMU,No_cycles,production_cycles,Production_day,month,weight_kg,number_treated,antibiotic_name,grp_api_used.concentartionofaai,grp_api_used.unitofaai,api_name,antibiotic_amount)%>% ###select relevant columns
  
  mutate(mass=grp_api_used.concentartionofaai*antibiotic_amount)%>% ###calculate the mass for each entry
  mutate(biomass=weight_kg*number_treated)%>%  ###calculate the biomass for each entry
  dplyr::mutate(total_mass=sum(mass,na.rm = TRUE))%>% ### sum mass for all entries
  dplyr::mutate(total_biomass=sum(biomass,na.rm = TRUE))%>%  ###sum biomass for all entries
  dplyr::mutate(total_mg_PCU=total_mass/total_biomass)%>%   ###calculate the overall mass-based metric (mg/PCU)
  dplyr::filter(!is.na(api_name)) #### remove entries/farms that did not use antibiotics



###calculate the total usage (mg/PCU) per antibiotic per farm per cycle

allfarmCycles2=allfarmCycles1%>%
  group_by(coversheet.farm_barcode,api_name,No_cycles)%>% ###group the the antibiotic per farm per cycle
  dplyr::mutate(total_mass_farm=sum(mass))%>% ##calculate the total mass per antibiotic in each farm per cycle
  dplyr::mutate(total_biomass_farm=sum(biomass))%>%  ##calculate the total biomass per antibiotic in each farm per cycle
  dplyr::ungroup()%>% ##ungroup the data for each farm
  dplyr::mutate(mg_pcu_farm=total_mass_farm/total_biomass_farm)%>% ###calculate the usage in mg/PCU (mass-based metric)
  dplyr::select(coversheet.farm_barcode,No_cycles,Farm_label,api_name,mg_pcu_farm)%>% ###select the relevant columns
  dplyr::distinct() ###select the unique entries only - removes duplicates
#####################################################################################

####get number of farms using each antibiotic during the study period in each cycle

allfarmCycles2_1<-allfarmCycles2%>%
  dplyr::group_by(api_name,No_cycles)%>% ###group antibiotic per cycle
  dplyr::mutate(no_farms=n_distinct(coversheet.farm_barcode))%>% ### count the number of farms using each antibiotic per cycle
  dplyr::ungroup()%>% ####ungroup the antibiotic per cycle
  dplyr::group_by(No_cycles)%>% ###group the data into cycles
  dplyr::mutate(median_mgPCU=median(mg_pcu_farm,na.rm = TRUE))%>% ###calculate the median usage in mg/PCU per cycle
  dplyr::ungroup() ###ungroup the cycles data


#### subset data to get data showing the median mg/PCU per cycle
median_mgPCU<-allfarmCycles2_1 %>%
  dplyr::select(No_cycles,median_mgPCU)%>%
  dplyr::distinct()


###plot a line graph to show the proportion of farms using an antibiotic in each cycle

cycleTrend<-allfarmCycles2_1%>%
  dplyr::mutate(no_farmsCycle=case_when(No_cycles==1~83,
                                        No_cycles==2~81,
                                        No_cycles==3~79,
                                        No_cycles==4~65,
                                        No_cycles==5~57,
                                        No_cycles==6~34))%>%  ####create a column with the total number of farms in each cycle
  dplyr::mutate(prop_farmsCycle=no_farms/no_farmsCycle*100)%>% ###calculate the proportion of farms using antimicrobial in each cycle
  dplyr::select(No_cycles,api_name,no_farms,no_farmsCycle,prop_farmsCycle)%>% ####select relevant columns
  dplyr::distinct() ###get only unique entries - no duplicates

 
####plot a line graph
 ggplot(cycleTrend,aes(x=No_cycles,y=prop_farmsCycle))+
    geom_line(linewidth=2.5,color="#CD5B45")+
    geom_point(size=4,color="#00CDCD")+
    facet_wrap(~api_name)+
    theme_bw()+
    labs(
      x="Number of production cycles",
      y= "Proportion of farms"
    )+
    scale_x_continuous(breaks = seq(1,6,1))+
    theme(
    axis.title = element_text(size = 20, face = "bold", colour = "black"),
    plot.title = element_text(size = 20, face = "bold", colour = "black", hjust = 0.5),
    axis.text = element_text(size = 20, face = "bold", color = "black"),
    legend.title = element_text(size = 22, face = "bold"),
    legend.text = element_text(size = 20, face = "bold"),
    strip.text.x = element_text(size = 20, colour = "black", face = "bold")
  )

#####save the line graph in the working folder
  
  tiff(filename = "AMU_cycle.tiff",type = "cairo", units = "mm",width = 540,height =340,pointsize = 20,res = 300)
  
ggplot(cycleTrend,aes(x=No_cycles,y=prop_farmsCycle))+
    geom_line(linewidth=3,color="#CD5B45")+
    geom_point(size=5,color="#00CDCD")+
    facet_wrap(~api_name)+
    theme_bw()+
    labs(
      x="Number of production cycles",
      y= "Proportion of farms"
    )+
    scale_x_continuous(breaks = seq(1,6,1))+
    theme(
    axis.title = element_text(size = 28, face = "bold", colour = "black"),
    plot.title = element_text(size = 28, face = "bold", colour = "black", hjust = 0.5),
    axis.text = element_text(size = 28, face = "bold", color = "black"),
    legend.title = element_text(size = 30, face = "bold"),
    legend.text = element_text(size = 28, face = "bold"),
    strip.text.x = element_text(size = 28, colour = "black", face = "bold")
  )

dev.off()


```



# Mortality analysis


```{r}

#####calculate mortality rate

##### calculate overall cycle mortality

cycle_mortality<-mortality%>%
  group_by(coversheet.farm_barcode,No_cycles)%>%  ##group mortality data per farm per cycle
  mutate(total_dead=sum(no.Dead,na.rm = TRUE))%>%  ##get the total number of dead chicken per cycle per farm
  ungroup()%>% ###ungroup the farm mortality per cycle
  dplyr::select(coversheet.farm_barcode,No_cycles,no_DOC,total_dead,AMU)%>% ###select relevant columns
  distinct()%>% ###subset unique data - no duplicates
  dplyr::mutate(mortality=total_dead/no_DOC*100)%>% ###calculate cycle mortality rate
  dplyr::mutate(cycles=case_when(
    No_cycles==1~"Cycle 1",
    No_cycles==2~"Cycle 2",
    No_cycles==3~"Cycle 3",
    No_cycles==4~"Cycle 4",
    No_cycles==5~"Cycle 5",
    No_cycles==6~"Cycle 6",
  ))  ###rename the number of cycles

  
  ### calculate the minimum, 1st quartile, median, mean, 3rd quartile and maximum cycle mortality per cycle
  summary(cycle_mortality$mortality[cycle_mortality$No_cycles==1])
  summary(cycle_mortality$mortality[cycle_mortality$No_cycles==2])
  summary(cycle_mortality$mortality[cycle_mortality$No_cycles==3])
  summary(cycle_mortality$mortality[cycle_mortality$No_cycles==4])
  summary(cycle_mortality$mortality[cycle_mortality$No_cycles==5])
  summary(cycle_mortality$mortality[cycle_mortality$No_cycles==6])

```


# Mixed-effects regression model


```{r}

######Run linear mixed-effects model to establish if dosing ratio is associated with daily mortality

dailyMortData<-daily_mortality%>%
  dplyr::select(coversheet.farm_barcode,No_cycles,Production_day,daily_mortality)###mortality data

daily_UDD_DDD1<-cycle1Metric_b%>%
  dplyr::select(coversheet.farm_barcode,No_cycles,Production_day,UDD_DDDTreat) ###dosing practices data

###combine mortality and dosing practices data
daily_mort_dose<-merge(daily_UDD_DDD1,dailyMortData,by.x = c("coversheet.farm_barcode","No_cycles","Production_day"),by.y =c("coversheet.farm_barcode","No_cycles","Production_day"),all = TRUE )%>%
  dplyr::filter(!is.na(UDD_DDDTreat))


###Run the linear mixed-effects model - daily mortality as fixed-effects and farm and cycle no. as random effects
dailyUseMort<-lmer(log(UDD_DDDTreat)~daily_mortality+ (1 | coversheet.farm_barcode/No_cycles), data = daily_mort_dose)

summary(dailyUseMort)

# Get predicted values for daily mortality and use (dosing practice) 
pred_mort <- ggpredict(dailyUseMort, terms = "daily_mortality")


###plot the predicted values
ggplot(pred_mort, aes(x, predicted)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2,fill = "steelblue") +
  theme_bw()+
  labs(x = "Daily mortality", y = "Predicted UDD/DDD dosing ratio",
       title = "Effect of daily mortality on UDD/DDD dosing ratio")+
  theme(
    axis.title = element_text(size = 20, face = "bold", colour = "black"),
    plot.title = element_text(size = 20, face = "bold", colour = "black"),
    axis.text = element_text(size = 20, face = "bold", color = "black"),
    legend.title = element_text(size = 22, face = "bold"),
    legend.text = element_text(size = 20, face = "bold"),
    strip.text.x = element_text(size=20, colour="Black", face="bold")
  )


```


####    Interpretation

This analysis quantified antimicrobial use using mass-based metrics (mg/PCU), explored temporal patterns across production cycles and evaluated association between mortality and dosing practices.

The workflow demonstrates how raw longitudinal surveillance data were transformed into analytical outputs suitable for epidemiological interpretation and decision-making.

#####   Reproducibility Notes

This workflow was developed using R and R Markdown.

The analytical process was organized into clearly defined stages including data cleaning, validation, metric calculation, regression modelling, visualization, and interpretation.

The original research data are not included because they are subject to data-sharing restrictions.



